---
title: "Building Metrics with gloBFPr"
output: html_document
---

# Analyze Building Footprints

This vignette shows how to compute building-level metrics from a building
footprint layer with height attributes. Most functions return the same `sf`
object with additional metric columns, so the usual workflow is to pipe or
reassign the result step by step.

```{r setup, message=FALSE}
library(gloBFPr)
library(sf)
library(dplyr)
```

## Example data

The package includes a small `sf` example dataset with building footprints,
unique IDs, and building heights.

```{r}
data(globfp_example)

buildings <- globfp_example
buildings <- buildings[seq_len(min(10, nrow(buildings))), ]

names(buildings)
```

For your own area of interest, first retrieve building footprints with
`search_3dglobdf()`, then pass the returned polygon layer to the metric
functions.

```{r eval=FALSE}
buildings <- search_3dglobdf(
  bbox = c(-83.065644, 42.333792, -83.045217, 42.346988),
  out_type = "poly",
  quiet = TRUE
)
```

## Morphological Metrics

`get_morphology()` computes geometric and shape metrics such as footprint area,
perimeter, volume, compactness, convexity, elongation, and pairwise distance.

```{r eval=FALSE}
morphology <- get_morphology(buildings, quiet = TRUE)

morphology |>
  st_drop_geometry() |>
  select(id, Height, g_area, pmeter, vol, rec, cnv, elo_z) |>
  head()
```

You can also request only a subset of metrics when you do not need the full
set.

```{r eval=FALSE}
basic_shape <- get_morphology(
  buildings,
  metrics = c("g_area", "pmeter", "vol", "rec"),
  quiet = TRUE
)
```

## Neighbor Metrics

`get_neighbors()` estimates the number of neighboring buildings and centroid
distance summaries using a fixed search radius and Voronoi adjacency.

```{r eval=FALSE}
neighbors <- get_neighbors(
  morphology,
  radius = 500,
  quiet = TRUE
)

neighbors |>
  st_drop_geometry() |>
  select(id, n_count, m_ndist, min_ndist, max_ndist, sd_ndist) |>
  head()
```

Use a smaller radius for dense local context or a larger radius for broader
neighborhood context.

```{r eval=FALSE}
neighbors_100m <- get_neighbors(buildings, radius = 100, quiet = TRUE)
```

## Population Density

`get_pop_density()` downloads GHSL population tiles and extracts total
population and population density at each building centroid. Set `distance` to
aggregate population within a buffer around each building centroid. This step
requires network access and can take time depending on the number of GHSL tiles
intersecting the study area.

```{r eval=FALSE}
population <- get_pop_density(
  neighbors,
  year = 2025,
  quiet = TRUE
)

population_500m <- get_pop_density(
  neighbors,
  year = 2025,
  distance = 500,
  quiet = TRUE
)

population |>
  st_drop_geometry() |>
  select(id, pop_total, pop_den) |>
  head()
```

## Residential Classification

`get_residential()` uses GHSL built-up surface data to estimate whether each
building is residential. The `threshold` is the minimum residential built-up
surface percentage in the matched GHSL cell.

```{r eval=FALSE}
residential <- get_residential(
  population,
  year = 2025,
  threshold = 80,
  quiet = TRUE
)

residential |>
  st_drop_geometry() |>
  select(id, res, res_pct, total_built_vals, res_vals, nres_vals) |>
  head()
```

## Distance to Nearest Greenspace

`get_dng()` computes the distance from each building centroid to the nearest
qualifying green patch within a buffer. The source can be `metachm`, `esri`, or
`sentinel2`.

```{r eval=FALSE}
dng <- get_dng(
  residential,
  datasource = "metachm",
  radius = 800,
  min_tree_height = 2,
  min_area = 500,
  unit = "m2",
  quiet = TRUE
)

dng |>
  st_drop_geometry() |>
  select(id, dng) |>
  head()
```

For a 2D greenery mask source, use `esri` or `sentinel2`.

```{r eval=FALSE}
dng_esri <- get_dng(
  buildings,
  datasource = "esri",
  zoom = 17,
  radius = 800,
  min_area = 500,
  unit = "m2",
  quiet = TRUE
)
```

## Building Green View Index

`get_bgvi()` estimates Building Green View Index from building viewpoints. The
DSM can include metaCHM canopy height, while the visible-green feature layer can
come from height-filtered canopy, 2D map-tile greenspace, or the union of both.
An OpenTopography API key is required for the DEM. This can be computationally
heavy; start with a small subset of buildings.

```{r eval=FALSE}
bgvi <- get_bgvi(
  buildings[1:3, ],
  datasource_canopy_height = "metachm",
  datasource_greenspace = "esri",
  min_tree_height = 2,
  radius = 800,
  floor = FALSE,
  workers = 1,
  key = "YOUR_OPENTOPOGRAPHY_API_KEY",
  quiet = TRUE
)

bgvi |>
  st_drop_geometry() |>
  select(id, mean_gvi, bottom_gvi, top_gvi) |>
  head()
```

To compute BGVI with a DSM built only from buildings and DEM, set
`datasource_canopy_height = NULL` and provide a 2D greenspace source.

```{r eval=FALSE}
bgvi_2d_green <- get_bgvi(
  buildings[1:3, ],
  datasource_canopy_height = NULL,
  datasource_greenspace = "esri",
  radius = 800,
  short_building_threshold = 6,
  workers = 1,
  key = "YOUR_OPENTOPOGRAPHY_API_KEY",
  quiet = TRUE
)
```

To compute directional BGVI, provide one or more directions and a field of view
in degrees. The function reuses each computed viewshed and filters visible
greenery by direction.

```{r eval=FALSE}
bgvi_directional <- get_bgvi(
  buildings[1:3, ],
  datasource_canopy_height = "metachm",
  datasource_greenspace = "esri",
  radius = 800,
  directions = c("southwest", "south", "southeast"),
  field_of_view = 45,
  workers = 1,
  key = "YOUR_OPENTOPOGRAPHY_API_KEY",
  quiet = TRUE
)

bgvi_directional |>
  st_drop_geometry() |>
  select(id, mean_gvi_south, gvi_bottom_south, gvi_top_south) |>
  head()
```

To compute GVI by floor, set `floor = TRUE`. Use `floor_step` to sample every
`n` floors and reduce runtime. The top estimated floor is always included.

```{r eval=FALSE}
bgvi_by_floor <- get_bgvi(
  buildings[1:3, ],
  datasource_canopy_height = "metachm",
  radius = 800,
  floor = TRUE,
  floor_step = 3,
  workers = 2,
  key = "YOUR_OPENTOPOGRAPHY_API_KEY",
  quiet = TRUE
)

bgvi_by_floor |>
  st_drop_geometry() |>
  select(id, estimated_floors, mean_gvi, bottom_gvi, top_gvi, min_gvi, max_gvi, sd_gvi) |>
  head()
```

## A Complete Metric Pipeline

For a typical analysis, build the result incrementally. Keep `quiet = TRUE` for
batch workflows, or set it to `FALSE` when you want messages and progress bars.

```{r eval=FALSE}
result <- buildings |>
  get_morphology(quiet = TRUE) |>
  get_neighbors(radius = 500, quiet = TRUE) |>
  get_pop_density(year = 2025, quiet = TRUE) |>
  get_residential(year = 2025, threshold = 80, quiet = TRUE) |>
  get_dng(
    datasource = "metachm",
    radius = 800,
    min_area = 500,
    unit = "m2",
    quiet = TRUE
  )

metric_table <- result |>
  st_drop_geometry() |>
  select(
    id, Height, g_area, vol, rec,
    n_count, m_ndist, min_ndist, max_ndist, sd_ndist,
    pop_total, pop_den, res, res_pct,
    dng
  )

head(metric_table)
```
