---
title: "gloBFPr_data_acquisition"
author: "Xiaohao Yang"
vignette: >
  %\VignetteIndexEntry{gloBFPr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## 1 Retrieve Building Footprints with Heights Using a Bounding Box

Set `out_type = "all"` o return a comprehensive list of outputs, including:
- `poly`: an `sf` object of building footprints,
- `binary`: a `terra` raster indicating presence/absence of buildings, and
- `graduated`: a `terra` raster representing building height.

Specify `cell_size = 1` to generate raster layers with 1-meter resolution, ensuring detailed spatial representation of building geometries within the defined area of interest.

```{r eval=FALSE}
buildings_list <- gloBFPr::search_3dglobdf(bbox = c(-83.065644,42.333792,-83.045217,42.346988),
                                           out_type = "all", 
                                           cell_size = 1)
```

Setting `mask = TRUE` ensures the height raster is masked by the building footprints.

```{r eval=FALSE}
buildings_rast <- gloBFPr::search_3dglobdf(bbox = c(-83.065644,42.333792,-83.045217,42.346988),
                                           out_type = "graduated_rast", 
                                           mask = TRUE, 
                                           cell_size = 1)
```

Setting `data_source = "GBA"` to get building data from GlobalBuildingAtlas.

```{r eval=FALSE}
buildings_gba <- gloBFPr::search_3dglobdf(bbox = c(-83.065644,42.333792,-83.045217,42.346988),
                                           out_type = "poly", 
                                           data_source = "GBA",
                                          keep_source_id=T)
```
