---
title: "gloBFPr"
author: "Xiaohao Yang"
vignette: >
  %\VignetteIndexEntry{gloBFPr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 1. Get Building Footprints

## Step 1 Get the Spatial Grid with the Metadata of the Dataset

```{r eval=FALSE}
metadata <- gloBFPr::get_metadata()
```

## Step 2.1 Retrieve Building Footprints with Heights Using a Bounding Box

Set `out_type = "all"` o return a comprehensive list of outputs, including:
- `poly`: an `sf` object of building footprints,
- `binary`: a `terra` raster indicating presence/absence of buildings, and
- `graduated`: a `terra` raster representing building height.

Specify `cell_size = 1` to generate raster layers with 1-meter resolution, ensuring detailed spatial representation of building geometries within the defined area of interest.

```{r eval=FALSE}
buildings_list <- gloBFPr::search_3dglobdf(bbox = c(-83.065644,42.333792,-83.045217,42.346988), 
                                           metadata = metadata, out_type = "all", cell_size = 1)
```

## Step 2.2 Retrieve Building Footprints with Heights Using a Bounding Box

Setting `mask = TRUE` ensures the height raster is masked by the building footprints.

```{r eval=FALSE}
buildings_list <- gloBFPr::search_3dglobdf(bbox = c(-83.065644,42.333792,-83.045217,42.346988), 
                                           metadata = metadata, out_type = "all", 
                                           mask = TRUE, cell_size = 1)
```

# 2. Analyze Building Footprints

Load example data in the package

```{r eval=FALSE}
data(globfp_example)
```

## Morphologic properties and indices

```{r eval=FALSE}
result <- gloBFPr::get_morphology(globfp_example)
```

## Population density

```{r eval=FALSE}
result <- gloBFPr::get_pop_density(result)
```

## Neighboring buildings

```{r eval=FALSE}
result <- gloBFPr::get_neighbors(result, radius = 200)
```

##

```{r eval=FALSE}
result <- gloBFPr::get_green_acc(result, 
                                 min_tree_height = 2, 
                                 radius = 500,
                                 key = '665a5ee08172bf9a8db62da803294c9f')
```
