---
title: "Building Shadows and Radiation with gloBFPr"
output: html_document
vignette: >
  %\VignetteIndexEntry{Building Shadows and Radiation with gloBFPr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Analyze Building Shadows and Radiation

This vignette demonstrates how to compute building shadow footprints, shadow
height, and surface radiation with `sf` and `terra` objects. The examples use a
small synthetic building layer (sampled in Detroit) so the vignette can run without downloading
external data.

## Workflow Summary

1. Prepare an `sf` building polygon layer with a numeric `Height` column.
2. Define `solar_time` and `time_zone` so the functions estimate solar
   position from the building-layer centroid.
3. Use `get_shadow_footprint()` for shadow polygons.
4. Use `get_shadow_height()` with `sf` points or a `terra::SpatRaster`.
5. Use `get_radiation()` to sample roof and facade radiation and visualize the
   returned `sf` layer.
   

```{r setup, message=FALSE}
library(gloBFPr)
library(sf)
library(terra)
```

## Example Buildings

The shadow and radiation functions expect an `sf` polygon layer with a numeric
height column. Building data returned by `search_3dglobdf()` already uses the
`Height` field, so the same default works here.

The package includes a small `sf` example dataset with building footprints,
unique IDs, and building heights.

```{r}
data(globfp_example)
data(globfp_example_dem)
data(globfp_example_canopy_height)

buildings <- globfp_example
# buildings <- buildings[seq_len(min(10, nrow(buildings))), ]
dem <- rast(globfp_example_dem)
canopy_height <- rast(globfp_example_canopy_height)

names(buildings)
```

For your own area of interest, first retrieve building footprints with
`search_3dglobdf()`, then pass the returned polygon layer to the metric
functions.

## Solar Position

Start with `solar_time` and `time_zone`. The functions estimate sun azimuth and
elevation from the time and the centroid of the building layer.

If you already have solar geometry from another source, you can use manual
`azimuth` and `elevation` values instead. Azimuth is measured clockwise from
north, and elevation is measured above the horizon.

```{r}
solar_time <- "2026-06-21 15:00:00"
time_zone <- "America/Detroit"

azimuth <- 135
elevation <- 35
```

## Shadow Footprints

`get_shadow_footprint()` returns an `sf` polygon layer containing one shadow
footprint for each building.

```{r}
shadow_footprints <- get_shadow_footprint(
  buildings,
  solar_time = solar_time,
  time_zone = time_zone,
  plot = TRUE,
  quiet = TRUE
)
```

To compare shadows from multiple times, pass a vector of `solar_time` values.
When `plot_overlap_gradient = TRUE`, the plot uses transparent gray shadows so
overlapping shadow areas appear darker.

```{r}
solar_times <- data.frame(
  sun_id = c("morning", "midday", "afternoon"),
  solar_time = c(
    "2026-06-21 09:00:00",
    "2026-06-21 12:00:00",
    "2026-06-21 16:00:00"
  )
)

multi_shadow_footprints <- get_shadow_footprint(
  buildings,
  solar_time = solar_times$solar_time,
  time_zone = time_zone,
  plot = TRUE,
  plot_overlap_gradient = TRUE,
  quiet = TRUE
)

unique(multi_shadow_footprints$sun_id)
```

Set `overlap_shadow = TRUE` when you want a dissolved cumulative shadow footprint. This
is useful for seeing the total ground area affected by all supplied solar
positions.

```{r}
combined_shadow_footprints <- get_shadow_footprint(
  buildings,
  solar_time = solar_times$solar_time,
  time_zone = time_zone,
  overlap_shadow = TRUE,
  plot = TRUE,
  quiet = TRUE
)
```

## Shadow Height at Points

Use an `sf` point layer when you need shadow height at specific locations.

```{r}
sample_points <- st_make_grid(
  st_as_sfc(st_bbox(shadow_footprints)),
  cellsize = 8,
  what = "centers"
)

sample_points <- st_sf(
  id = seq_along(sample_points),
  geometry = sample_points
)

shadow_height_points <- get_shadow_height(
  buildings,
  shadow_locations = sample_points,
  solar_time = solar_time,
  time_zone = time_zone,
  quiet = TRUE
)

sample_points$shadow_height <- shadow_height_points[, 1]

plot(st_geometry(shadow_footprints), col = "#E5E7EB", border = NA)
plot(st_geometry(buildings), col = "white", border = "#111827", add = TRUE)
plot(
  sample_points["shadow_height"],
  pch = 16,
  key.pos = 4,
  add = TRUE
)
```

## Shadow Height as a Terra Surface

Pass a `terra::SpatRaster` template to compute a gridded shadow-height surface.
If `shadow_locations = NULL`, `get_shadow_height()` creates a template automatically.

```{r}
template <- rast(
  xmin = st_bbox(shadow_footprints)[["xmin"]],
  xmax = st_bbox(shadow_footprints)[["xmax"]],
  ymin = st_bbox(shadow_footprints)[["ymin"]],
  ymax = st_bbox(shadow_footprints)[["ymax"]],
  resolution = 4,
  crs = st_crs(buildings)$wkt
)

shadow_height_surface <- get_shadow_height(
  buildings,
  shadow_locations = template,
  solar_time = solar_time,
  time_zone = time_zone,
  quiet = TRUE
)

plot(shadow_height_surface, main = "Shadow height")
plot(st_geometry(buildings), col = NA, border = "black", add = TRUE)
```

## Tree Canopy and Terrain

Tree canopy can be included as an additional shadow obstacle by passing a
canopy height map. If you also provide a DEM, the function compares canopy and
building shadows in absolute elevation and returns shadow heights above local
ground.

If you do not already have these rasters, the shadow and radiation functions can
retrieve them internally. Canopy height currently supports `datasource_canopy_height =
"metachm"` or `"ethCHM"`, and DEM retrieval requires an OpenTopography API key.

```{r eval=FALSE}
shadow_height_with_downloaded_trees <- get_shadow_height(
  buildings,
  shadow_locations = template,
  solar_time = solar_time,
  time_zone = time_zone,
  datasource_canopy_height = "metachm",
  key = "YOUR_OPENTOPOGRAPHY_API_KEY",
  min_tree_height = 2,
  quiet = TRUE
)

radiation_with_downloaded_trees <- get_radiation(
  buildings,
  solar_time = solar_time,
  time_zone = time_zone,
  solar_normal = 850,
  solar_diffuse = 120,
  datasource_canopy_height = "metachm",
  key = "YOUR_OPENTOPOGRAPHY_API_KEY",
  min_tree_height = 2,
  canopy_transmissivity = 0.15,
  quiet = TRUE
)
```

```{r}
plot(dem, main = "Sample DEM")
plot(st_geometry(buildings), col = NA, border = "black", add = TRUE)
plot(canopy_height, main = "Canopy height")
plot(st_geometry(buildings), col = NA, border = "black", add = TRUE)
```

The same `get_shadow_height()` call can now include tree canopy.

```{r}
shadow_footprints_with_trees <- get_shadow_footprint(
  buildings,
  solar_time = solar_time,
  time_zone = time_zone,
  canopy_height = canopy_height,
  dem = dem,
  min_tree_height = 1.5,
  quiet = TRUE,
  plot = TRUE
)
```

```{r}
shadow_height_with_trees <- get_shadow_height(
  buildings,
  shadow_locations = template,
  solar_time = solar_time,
  time_zone = time_zone,
  canopy_height = canopy_height,
  dem = dem,
  min_tree_height = 2,
  quiet = TRUE
)

plot(shadow_height_with_trees, main = "Building and canopy shadow height")
plot(st_geometry(buildings), col = NA, border = "black", add = TRUE)
```

## Building Surface Radiation

`get_radiation()` estimates radiation on a 3D `sf` surface grid. If no grid is
provided, the function samples roofs and facades from the building polygons.
`solar_normal` is direct normal irradiance and `solar_diffuse` is diffuse
horizontal irradiance.

```{r}
radiation <- get_radiation(
  buildings,
  solar_time = solar_time,
  time_zone = time_zone,
  solar_normal = 850,
  solar_diffuse = 120,
  grid_res = 8,
  plot = TRUE,
  quiet = TRUE
)

head(st_drop_geometry(radiation))
```

To include tree shade in the radiation estimate, pass the canopy height map and
DEM. `canopy_transmissivity` controls how much direct radiation passes through
tree shade; `0` is fully opaque and `1` has no effect.

```{r}
radiation_with_trees <- get_radiation(
  buildings,
  solar_time = solar_time,
  time_zone = time_zone,
  solar_normal = 850,
  solar_diffuse = 120,
  canopy_height = canopy_height,
  dem = dem,
  min_tree_height = 2,
  canopy_transmissivity = 0.15,
  grid_res = 8,
  plot = TRUE,
  quiet = TRUE
)

roof_building_only <- aggregate(
  total ~ building_id,
  data = st_drop_geometry(radiation[radiation$surface == "roof", ]),
  FUN = mean
)
roof_with_trees <- aggregate(
  total ~ building_id,
  data = st_drop_geometry(radiation_with_trees[radiation_with_trees$surface == "roof", ]),
  FUN = mean
)

roof_compare <- merge(
  roof_building_only,
  roof_with_trees,
  by = "building_id",
  suffixes = c("_building_only", "_with_trees")
)

head(roof_compare)
```

The returned object contains both roof and facade sample points. Plotting all
of them with the default `sf` color scale can be hard to read because the dense
facade points trace the building walls. A clearer first view is to show roof
radiation as colored points and draw facade points separately in a muted color.

```{r}
roof_radiation <- radiation[radiation$surface == "roof", ]
facade_radiation <- radiation[radiation$surface == "facade", ]

roof_cols <- hcl.colors(100, "YlOrRd", rev = TRUE)
roof_breaks <- cut(
  roof_radiation$total,
  breaks = 100,
  include.lowest = TRUE,
  labels = FALSE
)

plot(st_geometry(buildings), col = "grey95", border = "grey45")
plot(
  st_geometry(facade_radiation),
  pch = 16,
  cex = 0.25,
  col = "#2563EB55",
  add = TRUE
)
plot(
  st_geometry(roof_radiation),
  pch = 16,
  cex = 0.75,
  col = roof_cols[roof_breaks],
  add = TRUE
)
legend(
  "topright",
  legend = c("lower roof total", "higher roof total", "facade samples"),
  pch = 16,
  col = c(roof_cols[1], roof_cols[100], "#2563EB55"),
  bty = "n"
)
```

You can also inspect roofs and facades as separate maps.

```{r}
plot(roof_radiation["total"], pch = 16, cex = 0.8, key.pos = 4)
plot(st_geometry(buildings), col = NA, border = "grey30", add = TRUE)

plot(facade_radiation["total"], pch = 16, cex = 0.45, key.pos = 4)
plot(st_geometry(buildings), col = NA, border = "grey30", add = TRUE)
```

For multiple times, provide one irradiance value for each `solar_time`.

```{r}
solar_day <- c(
  "2026-06-21 08:00:00",
  "2026-06-21 11:00:00",
  "2026-06-21 14:00:00",
  "2026-06-21 17:00:00"
)

radiation_day <- get_radiation(
  buildings,
  solar_time = solar_day,
  time_zone = time_zone,
  solar_normal = c(500, 850, 900, 650),
  solar_diffuse = c(180, 120, 110, 150),
  grid_res = 8,
  quiet = TRUE
)

roof_day <- radiation_day[radiation_day$surface == "roof", ]
facade_day <- radiation_day[radiation_day$surface == "facade", ]

roof_day_breaks <- cut(
  roof_day$total,
  breaks = 100,
  include.lowest = TRUE,
  labels = FALSE
)

plot(st_geometry(buildings), col = "grey95", border = "grey45")
plot(
  st_geometry(facade_day),
  pch = 16,
  cex = 0.25,
  col = "#2563EB55",
  add = TRUE
)
plot(
  st_geometry(roof_day),
  pch = 16,
  cex = 0.75,
  col = roof_cols[roof_day_breaks],
  add = TRUE
)
legend(
  "topright",
  legend = c("lower roof total", "higher roof total", "facade samples"),
  pch = 16,
  col = c(roof_cols[1], roof_cols[100], "#2563EB55"),
  bty = "n"
)
```
